Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-09-14 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
Updated formatting for manuscript

---
title: "Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [book.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:  
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

### Summary
Predicting forest responses to drought is an increasingly critical task under climate change effects.
Part of the problem is due to the lack of studies analyzing the confluence of leaf hydraulic traits with biophysical parameters.
In this study, we analyze the interaction between these two trait groups using forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
Each drought scenario (1966, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
Individual-level growth responses to the three individual droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts. 
We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected.

@@

*First, what drives the observed tendency for large trees to suffer more during drought?*  
@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms (besides insects) for the observed greater drougth growth reductions of large trees: (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees. Counteracting these effects, (3) the larger root systems of larger trees may confer an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.

@@

*Second, how do species' traits - alone and in interaction with tree size - influence drought response?* 
Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests. The solution is a trait-based approach. Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and specific leaf area (SLA) ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).

@@

*Third, are responses similar or variable across individual drought years?*
Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.

@@

### Materials and Methods

*Study site*
Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78°08'43.4"W) [[Bourg et al. 2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013]. Dominant species include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems >= 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems >= 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2018 (**???**), were obtained through the ForestGEO data portal (www.forestgeo.si.edu).

![**Table 2. Summary of hypotheses, predictions and results** ](tables_figures/table_variables.png){width=500px}

@@

$$ DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right] $$

@@

**Table 3. Species analyzed here, listed in descending order of ANPP_stem. n cores and DBH range represented, and species traits** [*This replaces/combines the two remaining tables in this section. Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each crown position) species means for each trait]

```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/Table2.csv")
kable(table2, booktabs=TRUE, format = "latex", caption = "Overview of analyzed species, detailing DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements. Units of measurements are in mm (DBH), % (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

*Identification of drought years*
[**add description of climate data used in Fig. 1, NEON vertical profiles**]

@@

*Analysis* 
Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable, and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (cand_full). All variables with dAICc >2 were used as candidates for each scenario's best model (figure ???? (tested_traits_best))

### Results

@@

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: (resistance > 1.0): #% in 1964-66, #% un 1977, and #% in 1999. Within the context of mixed effects models, there were no significant differences in resistance across drougth years (Table 4).

@@

### Discussion
Discussion outline

@@

**misc content to integrate**
From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity. This paper mentions it would be good to have this data with respect to latent affects from drought.
### Acknowledgements

words

### Author Contribution ####
words

>>>>

---
title: "Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [book.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

### Summary
Predicting forest responses to drought is an increasingly critical task under climate change effects.
Part of the problem is due to the lack of studies analyzing the confluence of leaf hydraulic traits with biophysical parameters.
In this study, we analyze the interaction between these two trait groups using forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
Each drought scenario (1966, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
Individual-level growth responses to the three individual droughts  were stronger in three cases: taller trees in dominant canopy positions, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration). However, there was substantial variation in the best predictor variables across given droughts.
We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected. **Add discussion points**

@@

*First, what drives the observed tendency for large trees to suffer more during drought?*

@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms (besides insects) for the observed greater drougth growth reductions of large trees: (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees. Counteracting these effects, (3) the larger root systems of larger trees may confer an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.

@@

*Second, how do species' traits - alone and in interaction with tree size - influence drought response?*

Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests. The solution is a trait-based approach. Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and specific leaf area (SLA) ([Medeiros et al. 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).

@@

*Third, are responses similar or variable across individual drought years?*

Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.

@@

\clearpage

### Materials and Methods
*Study site*

Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [[Bourg et al. 2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013]. Dominant species include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

@@

Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems \>= 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998]. From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems \>= 10cm to analyze functional trait composition relative to tree height (all analyses described below). Census data, which were last updated in 2018 (**???**), were obtained through the ForestGEO [data portal](www.forestgeo.si.edu).

\begin{figure}[H]
![**Table 2. Summary of hypotheses, predictions and results** ](tables_figures/table_variables.png){width=500px}
\end{figure}

@@

$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$

@@

**Table 3. Species analyzed here, listed in descending order of ANPP_stem. n cores and DBH range represented, and species traits** [*This replaces/combines the two remaining tables in this section. Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each crown position) species means for each trait]
```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/Table2.csv")

#the percentage symbol has a double backslash because otherwise this writes out the latex code when making the table

kable(table2, booktabs=TRUE, caption = "Overview of analyzed species, detailing DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements. Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

\raggedright

*Identification of drought years*

[**add description of climate data used in Fig. 1, NEON vertical profiles**]

@@

*Analysis*

Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable, and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (cand_full). All variables with dAICc >2 were used as candidates for each scenario's best model (figure ???? (tested_traits_best))

\raggedright

\newpage
### Results

@@

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: (resistance > 1.0): #% in 1964-66, #% un 1977, and #% in 1999. Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).

@@

\newpage
### Discussion
Discussion outline

@@

**misc content to integrate**
From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity. This paper mentions it would be good to have this data with respect to latent affects from drought.
\newpage
### Acknowledgements
words

### Author Contribution 
words
